\systemEnding: Bootstrapping Domain-Specific Languages for Neurally-Guided Bayesian Program Learning

### 2018-02-09 paper.tex
more informative caption

Automatically inducing programs from examples is a long-standing goal
of artificial intelligence. Recent work has successfully used
symbolic search techniques (e.g., Metagol:~\cite{muggleton2015meta},
FlashFill:~\cite{gulwani2011automating}), neural networks trained from
a corpus of examples (e.g., RobustFill:~\cite{devlin2017robustfill},
DeepCoder:~\cite{balog2016deepcoder}), and hybrids of neural and
symbolic methods (e.g., Neural-guided deductive search~\cite{ngds}) to
synthesize programs for task domains such as string transformations,
list processing, and robot navigation and planning. However, all these
approaches -- symbolic, neural and neural-symbolic -- rely upon a
hand-engineered \emph{Domain-Specific Language} (DSL). The DSL is an
inventory of restricted programming primitives, encoding domain-specific
knowledge about the space of programs. In
practice
we often have only a few input/output examples for each
program to be induced, and thus success often hinges on having a good
DSL which
provides a crucial inductive bias for what would otherwise be an
unconstrained search through the space of all computable functions.
Here we ask, to what extent can we dispense with such highly
hand-engineered domain-specific languages?

@@

Viewed as an inference problem,
our goal is to infer the DSL (written $\mathcal{D}$)
along with a program (written $p$) for each set of inputs/outputs (\emph{tasks}, written $x$).
A DSL $\mathcal{D}$ is a collection of domain-specific subroutines.
We equip $\mathcal{D}$ with a real-valued weight vector $\theta$, and together
$(\mathcal{D},\theta)$ define a distribution over programs.
Fig.~\ref{graphicalModel}(a) diagrams this inference problem as a hierarchical Bayesian model.
Inference in this model is
difficult because the programs are unobserved,
and so we must solve a hard search problem to recover them. To make
search tractable we learn a bottom-up \emph{recognition
  model} (written $q(\cdot )$, Fig.~\ref{graphicalModel}(b)).
The recognition model $q$ is a neural
network that regresses from input/output pairs to a distribution over
programs likely to explain the input/outputs. $q(\cdot )$ implements an amortized
inference
scheme~\cite{le2016inference} for the generative %ritchie2016deep
model.

@@

What DSL does \system learn?
The learned DSL contains templates for polynomials of different orders,
which lets the algorithm quickly hone in on the kinds of functions that are most appropriate to this domain (Tbl.~\ref{regressionDSL}).
Examining the programs themselves,
one finds that the algorithm discovers representations for each of the polynomials that minimizes the number of continuous degrees of freedom.
For example, it represents the linear function $-x+2$ with the program
\code{(* (* (+ x }$\mathcal{R}$\code{) }$\mathcal{R}$\code{) x)} which has two continuous degrees of freedom, and represents quartic functions using the invented DSL primitive $f_6$ in Tbl.~\ref{regressionDSL}
which has five continuous degrees of freedom.
This arises from our Bayesian framing -- both the bias towards shorter programs and the likelihood model's BIC penalty.
\begin{table}\centering
\begin{tabular}{l}
  \toprule
  %% Primitives& $+,\times:\mathbb{R}\to \mathbb{R}\to \mathbb{R}$\\
  %% & $\mathcal{R}:\mathbb{R}$ (real valued parameter)\\\midrule
  %% Observation $x$& $N$ input/output examples: $\left\{(i_n,o_n) \right\}_{n\leq N}$\\
  %% \midrule
  %% Likelihood $\probability[x|p]$& $\propto\exp\left( - D\log N \right)\prod_{n\leq N}\mathcal{N}(p(i_n)|o_n)$
  %% \\\midrule
  $f_0\,=\,$\code{(+ $\mathcal{R}$)}\\
  $f_1($\code{x}$)\,=\,$\code{(* (* (+ x $\mathcal{R}$) $\mathcal{R}$) x)} \\
  $f_2($\code{x}$)\,=\,$\code{(+ $\mathcal{R}$ ($f_1$ x))} \\
  $f_3($\code{x}$)\,=\,$\code{(* x ($f_2$ x))}\\
  $f_4($\code{x}$)\,=\,$\code{($f_0$ ($f_3$ x))}\\
  $f_5($\code{x}$)\,=\,$\code{(* x ($f_4$ x))}\\
  $f_6($\code{x}$)\,=\,$\code{($f_0$ ($f_4$ x))} (\emph{4th order polynomial})
  \\\bottomrule
\end{tabular}
\caption{Some learned DSL primitives for symbolic regression. System starts out with addition, multiplication, and real numbers, and learns to build 4th order polynomials.}\label{regressionDSL}
\end{table}

@@

  \\\midrule\multicolumn{6}{c}{\emph{List functions (rich DSL)}}\\\midrule
  \% solved&\textbf{86\%} &84\% &60\%&74\%&70\%\\
  Solve time&  0.8s&0.7s&1.0s&1.0s&1.1s
  \\\bottomrule
  \end{tabular}
\caption{\% solved: fraction solved w/ 5 sec timeout. Solve time: averaged over solved tasks. MDL: $-\expect\left[\probability[x|p] \right]$. For domains other than symbolic regression MDL is 0 nats. For symbolic regression MDL is a proxy for \# of continuous degrees of freedom.}\label{baselineComparisons}  \end{table}

>>>>

Automatically inducing programs from examples is a long-standing goal
of artificial intelligence. Recent work has successfully used
symbolic search techniques (e.g., Metagol:~\cite{muggleton2015meta},
FlashFill:~\cite{gulwani2011automating}), neural networks trained from
a corpus of examples (e.g., RobustFill:~\cite{devlin2017robustfill}), and hybrids of neural and
symbolic methods (e.g., Neural-guided deductive search~\cite{ngds}, DeepCoder:~\cite{balog2016deepcoder}) to
synthesize programs for task domains such as string transformations,
list processing, and robot navigation and planning. However, all these
approaches -- symbolic, neural and neural-symbolic -- rely upon a
hand-engineered \emph{Domain-Specific Language} (DSL). The DSL is an
inventory of restricted programming primitives, encoding domain-specific
knowledge about the space of programs. In
practice
we often have only a few input/output examples for each
program to be induced, and thus success often hinges on having a good
DSL which
provides a crucial inductive bias for what would otherwise be an
unconstrained search through the space of all computable functions.
Here we ask, to what extent can we dispense with such highly
hand-engineered domain-specific languages?

@@

Viewed as a probabilistic inference problem,
our goal is to infer the DSL (written $\mathcal{D}$)
along with a program (written $p$) for each set of inputs/outputs (\emph{tasks}, written $x$).
A DSL $\mathcal{D}$ is a collection of domain-specific subroutines.
We equip $\mathcal{D}$ with a real-valued weight vector $\theta$, and together
$(\mathcal{D},\theta)$ define a distribution over programs.
Fig.~\ref{graphicalModel}(a) diagrams this inference problem as a hierarchical Bayesian model.
Inference in this model is
difficult because the programs are unobserved,
and so we must solve a hard search problem to recover them. To make
search tractable we learn a bottom-up \emph{recognition
  model} (written $q(\cdot )$, Fig.~\ref{graphicalModel}(b)).
The recognition model $q$ is a neural
network that regresses from input/output pairs to a distribution over
programs likely to explain the input/outputs. We can also view $q(\cdot )$ as implementing an amortized
inference
scheme~\cite{le2016inference}.

@@

What DSL does \system learn?
The learned DSL contains templates for polynomials of different orders,
which lets the algorithm quickly hone in on the kinds of functions that are most appropriate to this domain (Tbl.~\ref{regressionDSL}).
Examining the programs themselves,
one finds that the algorithm discovers representations for each of the polynomials that minimizes the number of continuous degrees of freedom.
For example, it represents the linear function $-x+2$ with the program
\code{(* (* (+ x }$\mathcal{R}$\code{) }$\mathcal{R}$\code{) x)} which has two continuous degrees of freedom, and represents quartic functions using the invented DSL primitive $f_6$ in Tbl.~\ref{regressionDSL}
which has five continuous degrees of freedom.
This arises from our Bayesian framing -- both the bias towards shorter programs and the likelihood model's BIC penalty.
\begin{table}\centering
\begin{tabular}{l}
  \toprule
  %% Primitives& $+,\times:\mathbb{R}\to \mathbb{R}\to \mathbb{R}$\\
  %% & $\mathcal{R}:\mathbb{R}$ (real valued parameter)\\\midrule
  %% Observation $x$& $N$ input/output examples: $\left\{(i_n,o_n) \right\}_{n\leq N}$\\
  %% \midrule
  %% Likelihood $\probability[x|p]$& $\propto\exp\left( - D\log N \right)\prod_{n\leq N}\mathcal{N}(p(i_n)|o_n)$
  %% \\\midrule
  $f_0\,=\,$\code{(+ $\mathcal{R}$)}\\
  $f_1($\code{x}$)\,=\,$\code{(* (* (+ x $\mathcal{R}$) $\mathcal{R}$) x)} \\
  $f_2($\code{x}$)\,=\,$\code{(+ $\mathcal{R}$ ($f_1$ x))} \\
  $f_3($\code{x}$)\,=\,$\code{(* x ($f_2$ x))}\\
  $f_4($\code{x}$)\,=\,$\code{($f_0$ ($f_3$ x))}\\
  $f_5($\code{x}$)\,=\,$\code{(* x ($f_4$ x))}\\
  $f_6($\code{x}$)\,=\,$\code{($f_0$ ($f_4$ x))} (\emph{4th order polynomial})
  \\\bottomrule
\end{tabular}
\caption{Some learned DSL primitives for symbolic regression. The system starts with addition, multiplication, and real numbers, and learns to build polynomials up to 4th order.}\label{regressionDSL}
\end{table}

@@

  \\\midrule\multicolumn{6}{c}{\emph{List functions (rich DSL)}}\\\midrule
  \% solved&\textbf{86\%} &84\% &60\%&74\%&70\%\\
  Solve time&  0.8s&0.7s&1.0s&1.0s&1.1s
  \\\bottomrule
  \end{tabular}
\caption{\% solved w/ 5 sec timeout. Solve time: averaged over solved tasks.
  RF/DC: trained like RobustFill/DeepCoder.
  PCFG: model w/o structure learning.
  Enum: model w/o learning.
  MDL: $-\expect\left[\probability[x|p] \right]$. For domains other than symbolic regression MDL is 0 nats. For symbolic regression MDL is a proxy for \# of continuous degrees of freedom.}\label{baselineComparisons}  \end{table}
